Cognate beginnings to bilingual lexical acquisition

### 2023-07-06 manuscript/manuscript.qmd
Fix appendix

---
title: "Cognate beginnings to bilingual lexical acquisition"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        city: Barcelona, 08005
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    email: avila.varela.daniela@gmail.com
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Linguistics Center (CLUL), School of Arts and Humanities
        city: Lisbon, 1600-214
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    email: jignaciocastillejo@gmail.com
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        city: Madrid, 28049
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    email: nuria.sebastian@upf.edu
    affiliations: 
      - ref: cbc
keywords:
  - cognate
  - word acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source
lang: en

format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: true
    doublespacing: false
    runninghead: "Cognate beginnings to bilingual lexical acquistion"
    include-in-header: 
      text: |
        \usepackage{tipa}
        \usepackage{booktabs}
abstract: |
    Bilingual infants show equivalent developmental trajectories of lexical acquisition compared to their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalise on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. GGC was supported by the This research was supported by a FPI 2020 research contract (PRE2019-088165) in a project funded by the Spanish Ministry of Science, Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported by the European Union's Horizon 2023 research and innovation program under Marie Skłodowska–Curie Grant (765556) and a postdoctoral fellowship from the Foundation for Science and Technology of Portugal (UIDB/00214/2020). NSG was supported by Academia 2019 award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
fig-dpi: 1000
---

@@

library(gt)
library(ggplot2)
library(patchwork)

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], that  are phonologically very similar. When the child is exposed to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos-"pe.ro}/ would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

@@

tbl_data <- items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(total = rowSums(across(a:d)),
           items = map(items, function(x) tail(x, 2))) |>  
    arrange(desc(total)) |> 
    relocate(semantic_category, a:d, total, items) |> 
    select(-total)

@@

sums_row <- as.data.frame(lapply(tbl_data, sum_rows),
                          col.names = colnames(tbl_data))
tbl_data <- rbind(tbl_data, sums_row)

@@

r_version <- glue::glue(r_info$R.version$major, 
                        r_info$R.version$minor, 
                        .sep = ".")
stan_version <- "2.32.1"

@@

We collected `r format(n_obs, big.mark = ",")` responses to `r format(nrow(n_items), big.mark = ",")` items. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest: *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (a language exposure-weighted lexical frequency), and *Cognateness* (defined as the phonological similarity between translation equivalents). We added these variables as main effects, together with the two-way and three-way interactions between *Age*, *Exposure*, and *Cognateness*. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [@barr2013random]. We specified a weakly informative prior around the parameters of the model. @eq-linear shows a detailed description of the model.

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}) \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\text{where:}\\
& u_{1-8_{i}}~\text{: participant-level adjustments} \\
& w_{1-3_{j}}~\text{: TE-level adjustments} \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2) \\
&\text{where }~\rho_{u_{0-8}, w_{0-3}}~\text{are the correlations between group-level adjustments}\\
\end{aligned}
$$ {#eq-linear}

@@

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-forms lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We then mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We then multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child whose degree of exposure is 80% for Catalan and 20% for Spanish, the expected *Exposure* score to the Catalan word *cavall* [*horse*]---with a lexical frequency of `r round(items$freq[items$item=="cat_cavall"], 2)`---would be `r round(items$freq[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq[items$item=="cat_cavall"]*0.20, 2)`.

@@

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome]. We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four iteration chains with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

post_draws_list <- model_summary |> 
    mutate(across(.median:.lower,
                  \(x) ifelse(grepl("Intercept", .type),
                              plogis(x), x/4))) |> 
    split(model_summary$.variable)
```
The model posterior showed adequate chain convergence diagnostics for all parameters ($\hat{R} \le 1.06$, $0.022 \le N_{eff}/N < 1.1001$) and little evidence of multicollinearity. @tbl-coefs shows the summary of the posterior distribution of the fixed regression coefficients, and their degree of overlap with the ROPE. For interpretability, we report the estimated regression coefficients transformed to the probability scale^[The logit and probability scales relate non-linearly. This means that one logit difference is not necessarily translated to a unique value in the probability scale. For example, the probability of acquisition of a given word might increase in 5% when age increases from 22 to 23 months, the probability of acquisition of the same word might only increase in 0.2% when age increases from 31 to 32 months. The linear growth of the probability of acquisition differs along the logistic curve, and therefore deciding the age point at which to report the estimates of the regression coefficients in the probability scale is not trivial. Following @gelman2020regression, we report the maximum value of such coefficient, which corresponds to the linear growth (i.e. derivative) of the logistic curve at the age at which most participants were acquiring a given word. This value can be approximated by dividing the coefficient in the logit scale by four: $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$.]. The resulting values correspond to the maximum difference in probability of acquisition (*Comprehension* or *Comprehension and Production*) that corresponds to a one standard deviation change in each predictor.

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients."
#| tbl-note: "Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. p(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Lines and error bands correspond to the mean and 95% credible interval of the posterior-predicted means. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are shown on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
fig_data <- model_epreds |> 
    summarise(.median = mean(.value),
              .lower = qi(.value)[,1],
              .upper = qi(.value)[,2],
              .by = c(ends_with("_std"), .category)) |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) 
fig_data |> 
    ggplot(aes(age, .median, 
               ymin = .lower,
               ymax = .upper,
               colour = lv, 
               fill = lv,
               linetype = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             linetype = "dotted") +
    annotate(geom = "text",
             x = min(responses$age),
             y = 1,
             hjust = 0,
             vjust = 0.9,
             label = "Observed\nage range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linewidth = 1/4,
               linetype = "dotted") +
    geom_hline(yintercept = 0.5, 
               linewidth = 1/4,
               linetype = "dotted") +
    geom_ribbon(alpha = 3/4,
                colour = NA) +
    geom_line(linewidth = 2/3) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)",
         linetype = "Cognateness (phonological similarity)") +
    scale_color_grey(start = 0.80, end = 0.1) +
    scale_fill_grey(start = 0.80, end = 0.1) +
    scale_linetype_manual(values = rev(c("solid", "dashed", "dotdash"))) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.box = "vertical",
          legend.title.align = 1,
          legend.justification = "right",
          legend.direction = "horizontal",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

@@

>>>>

---
abstract: |
    Bilingual infants show equivalent developmental trajectories of lexical acquisition compared to their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalise on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. GGC was supported by the a FPI 2020 research contract (PRE2019-088165) in a project funded by the Spanish Ministry of Science, Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported by the European Union's Horizon 2023 research and innovation program under Marie Skłodowska–Curie Grant (765556) and a postdoctoral fellowship from the Foundation for Science and Technology of Portugal (UIDB/00214/2020). NSG was supported by Academia 2019 award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
---

@@

library(knitr)
library(kableExtra)
library(ggplot2)

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [*cat*], that  are phonologically very similar. When the child is exposed to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos-"pe.ro}/ [*dog*] would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

@@

tbl_data <- items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(
        total = rowSums(across(a:d)),
        items = map(items, function(x) {
            paste(tail(x, 2),
                  collapse = ", ")
        })) |>  
    unnest_longer(items) |> 
    arrange(desc(total)) |> 
    relocate(semantic_category, a:d, total, items) |> 
    select(-total)

@@

sums_row <- tbl_data |> 
    summarise(semantic_category = "N",
              across(a:d, sum),
              items = "-")

tbl_data <- rbind(tbl_data, sums_row)

@@

r_version <- glue::glue(r_info$R.version$major, 
                        r_info$R.version$minor, 
                        .sep = ".")
stan_version <- "2.32.1"
```

@@

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome].

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest: *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (a language exposure-weighted lexical frequency), and *Cognateness* (defined as the phonological similarity between translation equivalents). We added these variables as main effects, together with the two-way and three-way interactions between *Age*, *Exposure*, and *Cognateness*. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [@barr2013random]. We specified a weakly informative prior around the parameters of the model. @eq-model shows a detailed description of the model.

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}) \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\text{where:}\\
& u_{1-8_{i}}~\text{: participant-level adjustments} \\
& w_{1-3_{j}}~\text{: TE-level adjustments} \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2) \\
&\text{where }~\rho_{u_{0-8}, w_{0-3}}~\text{are the correlations between group-level adjustments}\\
\end{aligned}
$$ {#eq-model}

@@

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-forms lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We then mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We then multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child whose degree of exposure is 80% for Catalan and 20% for Spanish, the expected *Exposure* score to the Catalan word *cotxe* [*car*]---with a lexical frequency of `r round(items$freq[items$item=="cotxe"], 2)`---would be `r round(items$freq[items$item=="cotxe"]*0.80, 2)`, while that of its translation to Spanish *coche* would be `r round(items$freq[items$item=="coche"]*0.20, 2)`.

@@

We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four iteration chains using the by-default No U-Turn Sampler algorithm with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

post_draws_list <- model_summary |> 
    mutate(across(.median:.lower,
                  \(x) ifelse(grepl("Intercept", .type),
                              plogis(x), x/4))) |> 
    split(model_summary$.variable)
```

Chain convergence diagnostics showed adequate values, and the model posterior showed little evidence of correlation, as indicated by negligible pairwise correlation between the marginal posterior distribution of coefficients. @tbl-coefs shows the summary of the posterior distribution of the fixed regression coefficients, and their degree of overlap with the ROPE. For interpretability, we report the estimated regression coefficients transformed to the probability scale^[The logit and probability scales relate non-linearly. This means that one logit difference is not necessarily translated to a unique value in the probability scale. For example, the probability of acquisition of a given word might increase in 5% when age increases from 22 to 23 months, the probability of acquisition of the same word might only increase in 0.2% when age increases from 31 to 32 months. The linear growth of the probability of acquisition differs along the logistic curve, and therefore deciding the age point at which to report the estimates of the regression coefficients in the probability scale is not trivial. Following @gelman2020regression, we report the maximum value of such coefficient, which corresponds to the linear growth (i.e. derivative) of the logistic curve at the age at which most participants were acquiring a given word. This value can be approximated by dividing the coefficient in the logit scale by four: $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$.]. The resulting values correspond to the maximum difference in probability of acquisition (*Comprehension* or *Comprehension and Production*) that corresponds to a one standard deviation change in each predictor.

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients. Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. p(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero. In the right column, we show a graphical representation of the median and 95% HDI of each coefficient, with a reference dotted line indicating the location of zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Lines and error bands correspond to the mean and 95% credible interval of the posterior-predicted means. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are shown on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
fig_data <- model_epreds |> 
    reframe(.median = mean(.value),
            .lower = qi(.value, .width = 0.95)[,1],
            .upper = qi(.value, .width = 0.95)[,2],
            .by = c(ends_with("_std"), .category)) |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) 
fig_data |> 
    ggplot(aes(age, .median, 
               ymin = .lower,
               ymax = .upper,
               colour = lv, 
               fill = lv,
               linetype = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             linetype = "dotted") +
    annotate(geom = "text",
             x = min(responses$age),
             y = 1,
             hjust = 0,
             vjust = 0.9,
             label = "Observed\nage range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linewidth = 1/4,
               linetype = "dotted") +
    geom_hline(yintercept = 0.5, 
               linewidth = 1/4,
               linetype = "dotted") +
    geom_lineribbon(aes(order = lv),
                    linewidth = 0.5) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)",
         linetype = "Cognateness (phonological similarity)") +
    scale_linetype_manual(values = rev(c("solid", "dashed", "dotdash"))) +
    scale_colour_grey(start = 0, end = 0) +
    scale_fill_grey(start = 0.85, end = 0.35) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.box = "vertical",
          legend.title.align = 1,
          legend.justification = "right",
          legend.direction = "horizontal",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

@@

# References
